|

### 2023-07-21 journal/03-methods.Rmd
minor edits; ready for review

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their team on defense, the opposing team on offense, and their assigned pass blocker.
We identify the pass blocker linked with the pass rusher of interest using the scouting data provided by PFF as mentioned in Section \ref{sec:data}.
Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the two player groups: pass rushers as $R$ and nearest pass blockers as $B$, as well as for the two team groups: defense $D$ and offense $O$.
We also account for attributes about pass rusher $j$ in play $i$ through the covariate vector $\mathbf{x_{ij}}$, and estimate their respective coefficients $\boldsymbol \beta$ as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation*}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + D_{d[i]} + O_{o[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
D_{d} &\sim N(\mu_D, \sigma^2_D), \text{ for } d = 1, \cdots, \text{ \# of defensive teams}, \\
O_{o} &\sim N(\mu_O, \sigma^2_O), \text{ for } o = 1, \cdots, \text{ \# of offensive teams}.
\end{aligned}
\end{equation*}

@@

As for the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects.
Our positional categorization for pass rushers include defensive end, defensive tackle, nose tackle, outside linebacker, interior linebacker (middle linebacker and inside linebacker), and Secondary (cornerback and safety).
As for pass blockers, we consider center, guard, tackle, and other (tight end, running back, fullback and wide receiver).
Note that these positions are encoded as indicator variables with defensive ends and tackles as the reference levels for the pass rushers and blockers, respectively.
We also account for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
Finally, we control for play-context covariates with respect to the offensive team.
These include the current down (first, second, third, fourth, or two-point conversion), yards to go for a first down, and current yardline (i.e., distance from the possession team's goal line). 
We consider play-context information since these variables impact a team's designed play, which may result in a play with low or high STRAIN regardless of the pass rusher's role. 
For instance, a team may call a short pass that is intended to be thrown early which could limit the amount of STRAIN on a play. 
Or a team may need to throw a deep pass which would require more time and potentially create more STRAIN.
We do not account for time directly in the model due to the concern that the time it takes for a quarterback to throw the ball is itself a function of both the play call and pressure from pass rushers.
Thus, since we do not know the designed play call, we condition on the play context to adjust for play-level differences attributing to a pass rusher's STRAIN.

>>>>

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their team on defense, the opposing team on offense, and their assigned pass blocker.
We identify the pass blocker linked with the pass rusher of interest using the scouting data provided by PFF as mentioned in Section \ref{sec:data}.
Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the two player groups: pass rushers as $R$ and nearest pass blockers as $B$, as well as for the two team groups: defense $D$ and offense $O$.
We also account for attributes about pass rusher $j$ in play $i$ through the covariate vector $\mathbf{x_{ij}}$, and estimate their respective coefficients $\boldsymbol \beta$ as fixed effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation*}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + D_{d[i]} + O_{o[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
D_{d} &\sim N(\mu_D, \sigma^2_D), \text{ for } d = 1, \cdots, \text{ \# of defensive teams}, \\
O_{o} &\sim N(\mu_O, \sigma^2_O), \text{ for } o = 1, \cdots, \text{ \# of offensive teams}.
\end{aligned}
\end{equation*}

@@

As for the fixed effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects.
Our positional categorization for pass rushers include defensive end, defensive tackle, nose tackle, outside linebacker, interior linebacker (middle linebacker and inside linebacker), and Secondary (cornerback and safety).
As for pass blockers, we consider center, guard, tackle, and other (tight end, running back, fullback and wide receiver).
Note that these positions are encoded as indicator variables with defensive ends and tackles as the reference levels for the pass rushers and blockers, respectively.
We also account for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
Finally, we control for play-context covariates with respect to the offensive team.
These include the current down (first, second, third, fourth, or two-point conversion), yards to go for a first down, and current yardline (i.e., distance from the possession team's goal line). 
We consider play-context information since these variables impact a team's designed play, which may result in a play with low or high STRAIN regardless of the pass rusher's role. 
For instance, a team may call a short pass that is intended to be thrown early which could limit the amount of STRAIN on a play. 
Or a team may need to throw a deep pass which would require more time and potentially create more STRAIN.
We do not account for time directly in the model due to the concern that the time it takes for a quarterback to throw the ball is itself a function of both the play call and pressure from pass rushers.
Thus, since we do not know the designed play call, we condition on the play context to adjust for play-level differences attributing to a pass rusher's STRAIN.

### 2023-07-21 journal/supplement.Rmd
minor edits; ready for review

\spacingset{1.45}

@@

Figure \ref{fig:fig_field} shows snapshots of a play from the week six game between the Las Vegas Raiders and Denver Broncos in the 2021 NFL regular season.
The play outcome is a sack made by Raiders defensive end Maxx Crosby on Broncos quarterback Teddy Bridgewater.
The locations of every Las Vegas (in black, on defense) and Denver (in orange, on offense) player on the field at 1, 2, 3, and 4 seconds after the ball snap is displayed, with Crosby highlighted in blue.
Note that the point size for each Las Vegas player corresponds to their estimated STRAIN as the play progresses.
This play is represented in Table 1 and Figure 1 (left) of the paper.

@@

```{r}
#| label: pass_rushers_count
#| fig.cap: "Distribution of the number of pass rushers on passing plays."
#| fig.width: 4
#| fig.height: 3
source(here("scripts", "12_fig_pass_rushers_count.R"))
pass_rushers_count
```

@@

We additionally consider a multilevel model to evaluate the impact of pass rushers on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
The average STRAIN by pass rusher $j$ on play $i$ is modeled as follows.
\begin{equation*}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation*}
This model version contains random intercepts for three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
As before, the coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
The variables included as fixed effects are the same as those in the model from Section 3.2 of the paper: position of pass rusher and blocker, number of blockers in a play, and play-context information (down, yards to go, and current yardline).

@@

We perform leave-one-week-out cross-validation to compare the multilevel models with and without team effects for play-level STRAIN.
For example, we train each model using data from weeks one through seven, and perform evaluation for week eight.
This procedure is repeated for every single holdout week, and we evaluate the model performance based on mean squared error (MSE).
Figure \ref{fig:fig_loocv_team_eff} shows the week-by-week holdout MSE for the aforementioned multilevel models.
We clearly see that a smaller MSE is achieved for each left-out week when accounting for team effects.
Furthermore, the model with team effects yields a lower overall holdout MSE ($4.1882$) than the one without random intercepts for teams ($4.4138$).
This provides a quantitative justification for including team effects in our multilevel model.

>>>>

\spacingset{1.5}

@@

Figure \ref{fig:fig_field} shows snapshots of a play from the week six game between the Las Vegas Raiders and Denver Broncos in the 2021 NFL regular season.
The play outcome is a sack made by Raiders defensive end Maxx Crosby on Broncos quarterback Teddy Bridgewater.
The locations of every Las Vegas (in black, on defense) and Denver (in orange, on offense) player on the field at 1, 2, 3, and 4 seconds after the ball snap is displayed, with Crosby highlighted in blue.
Note that the point size for each Las Vegas player corresponds to their estimated STRAIN as the play progresses.
This play is illustrated with Table 1 and Figure 1 (left) of the paper.

@@

```{r}
#| label: pass_rushers_count
#| fig.cap: "Distribution of the number of pass rushers on passing plays."
#| fig.width: 3.2
#| fig.height: 2.7
source(here("scripts", "12_fig_pass_rushers_count.R"))
pass_rushers_count
```

@@

We additionally consider a multilevel model to evaluate the impact of pass rushers on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
The average STRAIN by pass rusher $j$ on play $i$ is modeled as follows.
\begin{equation*}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation*}
This model version contains random intercepts for three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
As before, the coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed effects. 
The variables included as fixed effects are the same as those in the model from Section 3.2 of the paper, namely, position of pass rusher and blocker, number of blockers in a play, and play-context information (down, yards to go, and current yardline).

@@

We perform leave-one-week-out cross-validation to compare the multilevel models with and without team effects for play-level STRAIN.
For example, we train each model using data from weeks one through seven, and perform evaluation for week eight.
This procedure is repeated for every single holdout week, and we evaluate the model performance based on mean squared error (MSE).
Figure \ref{fig:fig_loocv_team_eff} shows the week-by-week holdout MSE for the aforementioned multilevel models.
We clearly see that a smaller MSE is achieved for each left-out week when accounting for team effects.
Furthermore, the model with team effects yields a lower overall holdout MSE ($4.1882$) than the one without random intercepts for teams ($4.4138$).
The out-of-sample performance justifies the inclusion of team effects in our multilevel model.
